Mechanization of Separation in Generic Extensions

### 2019-01-06 LICS/conclusion_m.tex
Conclusión, primera versión (casi final).

% Recordamos el objetivo del proyecto a "medio-término"
% (auto-ironía). Podemos luego mencionar la distinción entre thematic y
% programmatic (describiendo esta también). Luego mostramos ejemplos de
% decisiones en ambos aspectos: Thematic: renaming(?), locales para
% forcing (algo más?); Programmatic: trabajar en o, interfaz entre
% modelo (satisfacción de fórmulas internalizadas) y meta-satisfacción
% de axiomas relativizados en o. Indicando por qué esas decisiones nos
% parecieron adecuadas.

% Luego señalar los logros del paper poniéndolos en perspectiva del
% proyecto, destacando que sólo nos queda un axioma para tener M[G]|=
% ZF. (Alguien quizás se podría preguntar por qué no está C, si es así
% señalarlo). Explayarse sobre la prueba de replacement.

@@

\medskip
\fbox{Repite intro el p\'arrafo  que sigue}

\medskip

The proof of Separation depends on the Fundamental Theorems of
Forcing; in the spirit of the Formal Abstracts project we have only
stated those theorems in our current development, because from a
engineering point of view this can be done independently. The task of
proving the Fundamental Theorems is another landmark of our project,
after completing the formalization that generic extensions satisfy
the axiom scheme of Replacement.

@@

This statement is straightforwardly equivalent to that with $\Phi$
consisting of a single formula. The schematic nature of this result
hints at a proof by induction on formulas, and hence it must be shown
internally.

This is an appropriate point to insist that the internal/external
dichotomy has been a powerful agent in the shaping of our project.
This tension was also pondered by Paulson in his formalization of
G\"odel's constructible universe \cite{MR2051585}; after choosing a
shallow embedding of $\ZF$, every argument proved by induction on
formulas (or functions defined by recursion) should be done using
internalized formulas. Working on top of Paulson's library, we
prototyped the formula-transformer $\forceisa$, which is defined for
internalized formulas, and this affects indirectly the proof of the
Separation Axiom (despite the latter is not by induction). The need of
using internalized formulas reappears with the need of a general
version of the Reflection Principle (because the $\Phi$ there depends
on the instance of Replacement being proved).

A secondary outcome of this project is to assess which assumptions on
the ground model $M$ are needed to develop the forcing machinery. Up
to this point we have some anecdotal data; for example, to show that a
2-variable instance of Separation holds in $M[G]$, one needs to use a
6-variable instance in $M$. In the future, we expect to condense in a
new locale a finite list of (instances of) axioms on $M$ enough to
perform forcing constructions; this list will likely include all the
instances of Separation and Replacement in $M$ that are needed to
satisfy the requirements of the locales in the
\isatt{ZF-Constructible} session.

A few words are in order in relation to \isatt{ZF-Constructible}.
We believe that this  session  would benefit from some changes in
the way some results are organized accross theory files; for
intance, by cataloging all or most of the internalized formulas in one
file. Another, most basic example would be to start out with an even
more basic locale that only assumes $M$ to be a nonempty transitive class,
as many absoluteness results follow from these hypotheses.
As Paulson comments in the sources, it would have been better to
minimize the use of the Powerset Axiom in locales and proofs. There
are  useful natural models that satisfy a fraction of $\ZF$ not including
this particular axiom, and to ensure a broader applicability, it would
be convenient to have  absoluteness results not assuming it.
It is our desire to advocate a future work to a thorough
revision of the development of constructibility to address these needs
and to maximize modularity.

@@

>>>>

@@

% Repite intro el p\'arrafo  que sigue
%% The proof of Separation depends on the Fundamental Theorems of
%% Forcing; in the spirit of the Formal Abstracts project we have only
%% stated those theorems in our current development, because from a
%% engineering point of view this can be done independently. The task of
%% proving the Fundamental Theorems is another landmark of our project,
%% after completing the formalization that generic extensions satisfy
%% the axiom scheme of Replacement.

@@

% This statement is straightforwardly equivalent to that with $\Phi$
% consisting of a single formula. 
It is obvious that we can take the conjunction of $\Phi$ and state
Theorem~\ref{th:reflection-principle} for a single formula, say $\phi$.
The schematic nature of this result
hints at a proof by induction on formulas, and hence it must be shown
internally.

This is an appropriate point to insist that the internal/external
dichotomy has been a powerful agent in the shaping of our project.
This tension was also pondered by Paulson in his formalization of
G\"odel's constructible universe \cite{paulson_2003}; after choosing a
shallow embedding of $\ZF$, every argument proved by induction on
formulas (or functions defined by recursion) should be done using
internalized formulas. Working on top of Paulson's library, we
prototyped the formula-transformer $\forceisa$, which is defined for
internalized formulas, and this affects indirectly the proof of the
Separation Axiom (despite the latter is not by induction). The proof
of Replacement also calls for internalized formulas, because one needs
a general version of the Reflection Principle (since the formula
$\phi$ involved depends on the specific instance of Replacement being
proved).

A secondary, more prosaic, outcome of this project is to precisely
assess which assumptions on the ground model $M$ are needed to develop
the forcing machinery. The obvious are transitivity and $M$ being
countable (but keep in mind Lemma~\ref{lem:wf-model-implies-ctm}); the
first because many absoluteness results follows from this, the latter
for the existence of generic filters.  \fbox{Se me hace que la
  anécdota empieza con forces} \fbox{que agrega cuatro variables} A
more anecdotal one is that to show that any instance of Separation
with at most two parameters holds in $M[G]$, one needs to assume every
instance with six parameters in $M$ \fbox{Chequear los usos de any y
  every}. The purpose of identifying those assumptions is to assemble
in a locale the specific (instances of) axioms that should satisfy the
ground model in order to perform forcing constructions; this list will
likely include all the instances of Separation and Replacement that
are needed to satisfy the requirements of the locales in the
\isatt{ZF-Constructible} session.

We have already commented on our hacking of \isatt{ZF-Constructible}
to maximize its modularity and thus the re-usability in other
formalizations. We think it would be desirable to organize it
somewhat differently: a trivial change is to catalogue in one file all
the internalized formulas \fbox{No tiene la desventaja de que puede
  llegar a ser necesario} \fbox{ agregar más en un futuro.}. A more
conceptual modification would be to start out with an even more basic
locale that only assumes $M$ to be a non-empty transitive class, as
many absoluteness results follow from this hypothesis \fbox{Estaba en
  plural hipótesis, por qué?}. Furthermore, as Paulson comments in the
sources, it would have been better to minimize the use of the Powerset
Axiom in locales and proofs. There are useful natural models that
satisfy a sub-theory of $\ZF$ not including Powersete, and to ensure a
broader applicability, it would be convenient to have absoluteness
results not assuming it. Once we gained more confidence that these
ideas, we plan to contribute back to the official distribution of
Isabelle/ZF with a thorough revision of the development of
constructibility.

@@

The second author is grateful to the others two. They are very nice persons.
